---
title: "Pre- and post-20mph models models for Glasgow single-vehicle collisions"
output: html_notebook
date: 03/2025
author: Torran Semple (torran.semple@nottingham.ac.uk or torranas@gmail.com)
---

<br>

#### Notebook description:
This notebook describes the modelling process for the Glasgow models in the manuscript: 'Twelve Years of Evidence: Modelling the Injury Severity of Single-Vehicle Collisions pre- and post-20mph Implementation in Edinburgh and Glasgow' 

<br>

#### Note:
The data used in this notebook were derived from the UK Gov.'s STATS19 database (see: https://www.data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-accidents-safety-data)

<br>

## Overview of contents (Section 1: Glasgow pre-20mph)
1.1  Loading and viewing single-vehicle pre-20mph Glasgow Road Traffic Collisions (RTCs) (`GLA_SV_PRE`)
<br>
1.2  Check and generate independent variables for `GLA_SV_PRE` (e.g., combined age bands)
<br>
1.3  Correlation matrix for potential independent variables in `GLA_SV_PRE` (to control for multicollinearity)
<br>
1.4  Recoding, reordering & examining the distribution of the `GLA_SV_PRE` dependent variable (`accident_severity`)
<br>
1.5  Stepwise regression for `GLA_SV_PRE` using the `MASS` package: identifying Fixed Parameters Ordered Probit (FPOP) model starting variables
<br>
1.6  Specification of more complex models (e.g., Random Parameters Ordered Probit (RPOP) and Random Parameters Ordered Probit models with Heterogeneity in the Means (RPOPHM)) using the `Rchoice` package
<br>
1.7  Model evaluation

<br>
<br>

## Overview of contents (Section 2: Glasgow post-20mph)
This section mirrors the steps in Section 1, using the Glasgow single-vehicle post-20mph data (`GLA_SV_POST`); i.e., 2.1 loading and viewing data, 2.2 check and generate IVs etc...

<br>
<br>
<br>

## Set working directory, load/install required packages and package dependencies
```{r, warning=FALSE}

# Note: set the working directory to the source file location; for example:
setwd("~/Documents/collision_severity_analysis/Glasgow")

# Install required packages if necessary
if(!require(car)) install.packages("car")
if(!require(MASS)) install.packages("MASS")
if(!require(Rchoice)) install.packages("Rchoice")
if(!require(lmtest)) install.packages("lmtest")


# Load required packages
library(car) # For variable recoding, regression diagnostics and analysis
library(MASS) # For ordered regression
library(Rchoice) # For choice modeling
library(lmtest) # For likelihood-ratio tests


# Dependencies - shows R and package versions used in original analysis
sessioninfo::session_info()


```

<br>
<br>
<br>

# -----------------------------------------------------------------------------
# Section 1: Glasgow pre-20mph models
# -----------------------------------------------------------------------------

<br>
<br>

## 1.1 Loading and viewing GLA_SV_PRE
```{r}

# Load the Edinburgh pre-20mph data (GLA_SV_PRE) from GitHub (stored in 'collision_severity_analysis' repo) (alternatively, load 'GLA_SV_PRE.csv' locally)
GLA_SV_PRE <- read.csv("https://raw.githubusercontent.com/torransemple/collision_severity_analysis/refs/heads/main/collision_severity_analysis/Glasgow/data/GLA_SV_PRE20_model_data.csv", header=TRUE, sep = ",")

# Omit NAs, only complete observations can be used (sample reduces from 2505 to 2332 observations)
GLA_SV_PRE <- na.omit(GLA_SV_PRE) 

# Check structure of GLA_SV_PRE - all available variables, types etc. 
str(GLA_SV_PRE)

# Drop irrelevant columns (inc. duplicates of variables etc.)
GLA_SV_PRE <- GLA_SV_PRE[ -c(1:2, 4:47) ]

```


## 1.2 Independent variable inclusion criteria for GLA_SV_PRE
```{r}

# Note: potential independent variables must have at least 2% indicator group or no more than 98%

# Set minimum and maximum thresholds
min_threshold <- 0.02
max_threshold <- 0.98

# Calculate mean indicator group per column (i.e., independent variable)
col_means <- colMeans(GLA_SV_PRE, na.rm = TRUE)

# Find columns outside the threshold range (<0.02 or >0.98)
outlier_cols <- names(col_means[col_means < min_threshold | col_means > max_threshold])

# Print the results
if (length(outlier_cols) > 0) {
  message("Columns with mean outside", min_threshold, "and", max_threshold, ":")
  print(outlier_cols)
} else {
  message("No columns found with mean outside the specified range.")
}

# Hence, do not include variables listed below (interestingly this includes speed_limit_20 for GLA_SV_PRE, i.e., less than 2% of accidents occurred on 20mph roads)

```


## 1.2 continued: aggregation of age-related IVs
```{r}

# Some potential independent variables, especially those related to the age of the main casualty/driver, should be aggregated to more intuitive intervals as follows

# Create new variable for age_band_casualty_under_10
new_var_name <- "age_band_of_casualty_cas1_under_10"

# Create the new variable using ifelse
GLA_SV_PRE[new_var_name] <- rowSums(GLA_SV_PRE[, c("age_band_of_casualty_cas1_0_5", "age_band_of_casualty_cas1_6_10")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_PRE[, "age_band_of_casualty_cas1_under_10"] <- as.numeric(GLA_SV_PRE[, "age_band_of_casualty_cas1_under_10"])



# Create new variable for age_band_casualty_11_20
new_var_name <- "age_band_of_casualty_cas1_11_20"

# Create the new variable using ifelse
GLA_SV_PRE[new_var_name] <- rowSums(GLA_SV_PRE[, c("age_band_of_casualty_cas1_11_15", "age_band_of_casualty_cas1_16_20")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_PRE[, "age_band_of_casualty_cas1_11_20"] <- as.numeric(GLA_SV_PRE[, "age_band_of_casualty_cas1_11_20"])



# Create new variable for age_band_casualty_over_65
new_var_name <- "age_band_of_casualty_cas1_over_65"

# Create the new variable using ifelse
GLA_SV_PRE[new_var_name] <- rowSums(GLA_SV_PRE[, c("age_band_of_casualty_cas1_66_75", "age_band_of_casualty_cas1_over_75")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_PRE[, "age_band_of_casualty_cas1_over_65"] <- as.numeric(GLA_SV_PRE[, "age_band_of_casualty_cas1_over_65"])



# Create new variable for age_band_of_driver_veh1_over_65
new_var_name <- "age_band_of_driver_veh1_over_65"

# Create the new variable using ifelse
GLA_SV_PRE[new_var_name] <- rowSums(GLA_SV_PRE[, c("age_band_of_driver_veh1_66_75", "age_band_of_driver_veh1_over_75")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_PRE[, "age_band_of_driver_veh1_over_65"] <- as.numeric(GLA_SV_PRE[, "age_band_of_driver_veh1_over_65"])



# Create new variable for age_band_of_driver_veh1_16_25
new_var_name <- "age_band_of_driver_veh1_16_25"

# Create the new variable using ifelse
GLA_SV_PRE[new_var_name] <- rowSums(GLA_SV_PRE[, c("age_band_of_driver_veh1_16_20", "age_band_of_driver_veh1_21_25")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_PRE[, "age_band_of_driver_veh1_16_25"] <- as.numeric(GLA_SV_PRE[, "age_band_of_driver_veh1_16_25"])


```


## 1.3 Correlation matrix for potential IVs in GLA_SV_PRE models
```{r, warning=FALSE}

# To control for multicollinearity between potential independent variables, those with pairwise correlation coefficients >0.3 (or <-0.3) were considered too highly correlated to be included in the same model

# Create a correlation matrix for all variables in GLA_SV_PRE
cor_matrix <- cor(GLA_SV_PRE)

# Write csv to inspect correlation matrix in a more intuitive environment; note, it is often beneficial to apply coloured conditional formatting in Excel to identify variables that exceed 0.3 or -0.3
write.csv(x=cor_matrix, file="Cor_GLA_SV_PRE.csv")

```


## 1.4 Recoding, reordering & examining the distribution of the GLA_SV_PRE dependent variable (accident severity)
```{r}

# Prior to modelling, the DV ('accident_severity') had to be recoded as a factor; further, the order of 'accident_severity' was inverted so that 1=slight, 2=serious and 3=fatal

# recode DV (using 'car' package) so that 3 is highest outcome, i.e., most severe accidents
GLA_SV_PRE$accident_severity <- car::recode(GLA_SV_PRE$accident_severity, as.factor=TRUE, "3=1; 2=2; 1=3") 

# Check distribution of DV - 3 should have the fewest outcomes, 1 the most - i.e., heavily zero-inflated distribution
table(GLA_SV_PRE$accident_severity) 

# Check % of fatal RTCs (should be 1% or more for ordered modelling, or above 20 observations)
fatal <- (40/2332)*100
fatal

# Note: the % of fatal RTCs is 1.7% and >20; hence, ordered models are appropriate for GLA_SV_PRE

```


## 1.5 Stepwise regression for GLA_SV_PRE using the 'MASS' package
```{r}

# A stepwise regression (forward selection) algorithm was used to select the starting variables in the fixed parameters ordered probit (FPOP) framework

# Specify model start point with constant (1) only
FitStart = polr(accident_severity ~ 1, data=GLA_SV_PRE, na.action = na.omit, method = "probit") 

# Note: the following variables cannot be included due to insufficieny indicator group: "first_road_class_C", "speed_limit_20", "speed_limit_30", "vehicle_type_veh1_cycle", "bus_passenger_cas1_boarding", "bus_passenger_cas1_alighting", "bus_passenger_cas1_standing", "bus_passenger_cas1_seated", "casualty_type_cas1_cyclist", "casualty_type_cas1_taxi", "casualty_type_cas1_motorcylces"

# Specify potential full model for stepwise regression (checked against correlation matrix; also some variables, including propulsion code, junction control and engine capacity were removed due to sparsity or lack of variation)
FitAll = polr(accident_severity ~ casualty_type_cas1_pedestrian +
                pedestrian_location_cas1_footway_verge + 
                pedestrian_location_cas1_in_carriageway_not_crossing + 
                pedestrian_location_cas1_not_using_crossing + 
                first_road_class_A + first_road_class_B +
                road_type_one_way + 
                road_type_dual_c + 
                junction_detail_crossroads + 
                pedestrian_crossing_physical_facilities_pelican_puffin +
                pedestrian_crossing_physical_facilities_pedestrian_phase + 
                light_conditions_dark_lit + 
                weather_conditions_rain_no_wind + 
                vehicle_manoeuvre_veh1_reversing +
                vehicle_manoeuvre_veh1_slowing_stopping + 
                vehicle_manoeuvre_veh1_going_ahead_other + 
                first_point_of_impact_veh1_back + 
                first_point_of_impact_veh1_offside + 
                first_point_of_impact_veh1_nearside + 
                journey_purpose_of_driver_veh1_commute + 
                sex_of_driver_veh1_male + 
                age_band_of_driver_veh1_16_25 + 
                age_band_of_driver_veh1_over_65 + 
                age_band_of_driver_veh1_over_75 + 
                sex_of_casualty_cas1_male + 
                age_band_of_casualty_cas1_under_10 + 
                age_band_of_casualty_cas1_11_20 + 
                age_band_of_casualty_cas1_36_45 + 
                age_band_of_casualty_cas1_over_65, data=GLA_SV_PRE, na.action = na.omit, method = "probit")

# Forward selection algorithm (iterative process based on largest reduction in AIC per potential IV added to model formula)
step(FitStart, direction="forward", scope=formula(FitAll)) 


```


## 1.6 The specification of more complex models (e.g. RPOP or RPOPHM) using the 'Rchoice' package, for GLA_SV_PRE data
```{r}

# The 'Rchoice' package allows the specification of more complex models (i.e., RPOP and RPOPHM)

# Re-estimate basic FPOP model (i.e., FPOP_GLA_SV_PRE) based on the output of the previous chunk (i.e., the stepwise regression)
# Note only statistically significant IVs are retained: Pr>|z| (i.e., p-value) < 0.05 (95%) or <0.1 (90%)
FPOP_GLA_SV_PRE <- Rchoice(accident_severity ~ age_band_of_casualty_cas1_over_65 + 
                             vehicle_manoeuvre_veh1_going_ahead_other + 
                             casualty_type_cas1_pedestrian + 
                             sex_of_driver_veh1_male + 
                             light_conditions_dark_lit + 
                             first_point_of_impact_veh1_offside + 
                             first_road_class_A, data=GLA_SV_PRE, family=ordinal("probit"))

# Summarise FPOP model
summary(FPOP_GLA_SV_PRE)

```


## 1.6 continued: random parameter trials
```{r}

# Now trial random parameters (using the 'Trial_RP_GLA_SV_PRE' formula); note: this is a trial and error process, i.e., try every potential IV (that meets the previously defined inclusion criteria as random parameters)

# Trial RPOP 
Trial_RP_GLA_SV_PRE <- Rchoice(accident_severity ~ age_band_of_casualty_cas1_over_65 +
                                 vehicle_manoeuvre_veh1_going_ahead_other + 
                                 casualty_type_cas1_pedestrian +
                                 sex_of_driver_veh1_male + 
                                 light_conditions_dark_lit + 
                                 first_point_of_impact_veh1_offside + 
                                 first_road_class_A + 
                                 road_surface_conditions_wet + 
                                 pedestrian_location_cas1_footway_verge + 
                                 road_type_one_way, data=GLA_SV_PRE, family=ordinal("probit"),
                                 ranp=c(age_band_of_casualty_cas1_over_65="n", # specify random parameters (ranp); n = normal distribution
                                 road_surface_conditions_wet="n", 
                                 pedestrian_location_cas1_footway_verge="n",
                                 road_type_one_way="n"), 
                                 R=200) # R = number of Halton draws - 200 is sufficient for ranp trials

# build on FPOP model - trial all available independent variables as random parameters (around 6/7 variables specified as RPs at a time to reduce model estimation time



# Finalised RPOP model
RP_GLA_SV_PRE <- Rchoice(accident_severity ~ age_band_of_casualty_cas1_over_65 + 
                           vehicle_manoeuvre_veh1_going_ahead_other + 
                           casualty_type_cas1_pedestrian + 
                           sex_of_driver_veh1_male + 
                           light_conditions_dark_lit + 
                           first_point_of_impact_veh1_offside + 
                           first_road_class_A + road_surface_conditions_wet + 
                           pedestrian_location_cas1_footway_verge, data=GLA_SV_PRE, family=ordinal("probit"),
                           ranp=c(age_band_of_casualty_cas1_over_65="n", # specify random parameters (ranp); n = normal distribution
                           road_surface_conditions_wet="n", 
                           pedestrian_location_cas1_footway_verge="n"), 
                           R=500) # R = number of Halton draws - 500 required for final model

# Summarise RPOP model
summary(RP_GLA_SV_PRE)


```


## 1.6 continued: re-estimate FPOP with RP variables (for purposes of fair comparison of FPOP and RPOP in model evaluation)
```{r}

# Finalised FPOP model
FP_GLA_SV_PRE <- Rchoice(accident_severity ~ age_band_of_casualty_cas1_over_65 + 
                           vehicle_manoeuvre_veh1_going_ahead_other + 
                           casualty_type_cas1_pedestrian + 
                           sex_of_driver_veh1_male + 
                           light_conditions_dark_lit + 
                           first_point_of_impact_veh1_offside + 
                           first_road_class_A + road_surface_conditions_wet + 
                           pedestrian_location_cas1_footway_verge, data=GLA_SV_PRE, family=ordinal("probit"))

# Summarise FPOP model 
summary(FP_GLA_SV_PRE)

```


## 1.6 continued: test for heterogeneity in the means of random parameter variables (i.e., RPOPHM trials)
```{r}

# Random Parameters Ordered Probit with Heterogeneity in the Means of RPs (RPOPHM) trials can be conducted as shown in 'Edinburgh_models.Rmd'; In the case of GLA_SV_PRE, the RPOPHM framework produced statistically significant results (i.e., exogenous variables explained heterogeneity in the means of random parameter variables)

# Finalised RPOPHM model
RPHM_GLA_SV_PRE <- Rchoice(accident_severity ~ age_band_of_casualty_cas1_over_65 + 
                             vehicle_manoeuvre_veh1_going_ahead_other + 
                             casualty_type_cas1_pedestrian + 
                             sex_of_driver_veh1_male + 
                             light_conditions_dark_lit + 
                             first_point_of_impact_veh1_offside + 
                             first_road_class_A + 
                             road_surface_conditions_wet + 
                             pedestrian_location_cas1_footway_verge | age_band_of_casualty_cas1_over_65 + 
                             road_surface_conditions_wet + 
                             pedestrian_location_cas1_footway_verge + 
                             first_point_of_impact_veh1_back, data=GLA_SV_PRE, family=ordinal("probit"),
                             ranp=c(age_band_of_casualty_cas1_over_65="n", # specify random parameters (ranp); n = normal distribution
                             road_surface_conditions_wet="n", 
                             pedestrian_location_cas1_footway_verge="n"), 
                             R=500, # R = number of Halton draws - 500 required for final model
                             mvar=list( pedestrian_location_cas1_footway_verge=c("first_point_of_impact_veh1_back"))) # Specification of HM in RP variable(s)

# Summarise RPOPHM model
summary(RPHM_GLA_SV_PRE)

```


## 1.7: Model evaluation
```{r}

# Likelihood ratio tests are utilised in this analysis (using 'lmtest') to compare the final RPOP model (in this case RP_GLA_SV_PRE) to the fixed parameters counterpart (FP_GLA_SV_PRE) (LRT I); and also, the final RPOPHM model to the FPOP counterpart (LRT II)

# LRTI: LRT test for RP_GLA_SV_PRE versus FP_GLA_SV_PRE
LRTI <- lrtest(FP_GLA_SV_PRE, RP_GLA_SV_PRE)
LRTI

# The LRTI result shows that that RP_GLA_SV_PRE is superior to FP_GLA_SV_PRE with >99% level of confidence


# LRTII: LRT test for RPHM_GLA_SV_PRE versus FP_GLA_SV_PRE
LRTII <- lrtest(FP_GLA_SV_PRE, RPHM_GLA_SV_PRE)
LRTII

# The LRTII result shows that that RPHM_GLA_SV_PRE is superior to FP_GLA_SV_PRE with >99% level of confidence


```


## -----------------------------------------------------------------------------
## Final model for GLA_SV_PRE = RPOPHM (RPHM_GLA_SV_PRE)
## -----------------------------------------------------------------------------




<br>
<br>
<br>
<br>
<br>

# -----------------------------------------------------------------------------
# Section 2: Glasgow post-20mph models
# -----------------------------------------------------------------------------

<br>
<br>

## 2.1 Loading and viewing GLA_SV_POST
```{r}

# Load the Edinburgh pre-20mph data (GLA_SV_POST) from GitHub (stored in 'collision_severity_analysis' repo) (alternatively, load 'GLA_SV_POST.csv' using your PC's pathname)
GLA_SV_POST <- read.csv("https://raw.githubusercontent.com/torransemple/collision_severity_analysis/refs/heads/main/collision_severity_analysis/Glasgow/data/GLA_SV_POST20_model_data.csv", header=TRUE, sep = ",")

# Omit NAs, only complete observations can be used (sample reduces from 1590 to 1373 observations)
GLA_SV_POST <- na.omit(GLA_SV_POST) 

# Check structure of GLA_SV_POST - all available variables, types etc. 
str(GLA_SV_POST)

# Drop irrelevant columns (inc. duplicates of variables etc.)
GLA_SV_POST <- GLA_SV_POST[ -c(1:2, 4:6, 8:47) ]

```


## 2.2 Independent variable inclusion criteria for GLA_SV_POST
```{r}

# Note: potential independent variables must have at least 2% indicator group or no more than 98%

# Set minimum and maximum thresholds
min_threshold <- 0.02
max_threshold <- 0.98

# Calculate mean indicator group per column (i.e., independent variable)
col_means <- colMeans(GLA_SV_POST, na.rm = TRUE)

# Find columns outside the threshold range (<0.02 or >0.98)
outlier_cols <- names(col_means[col_means < min_threshold | col_means > max_threshold])

# Print the results
if (length(outlier_cols) > 0) {
  message("Columns with mean outside", min_threshold, "and", max_threshold, ":")
  print(outlier_cols)
} else {
  message("No columns found with mean outside the specified range.")
}

# Hence, do not include variables listed below

```

## 2.2 continued: aggregation of age-related IVs
```{r}

# Some potential independent variables, especially those related to the age of the main casualty/driver, should be aggregated to more intuitive intervals as follows

# Create new variable for age_band_casualty_under_10
new_var_name <- "age_band_of_casualty_cas1_under_10"

# Create the new variable using ifelse
GLA_SV_POST[new_var_name] <- rowSums(GLA_SV_POST[, c("age_band_of_casualty_cas1_0_5", "age_band_of_casualty_cas1_6_10")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_POST[, "age_band_of_casualty_cas1_under_10"] <- as.numeric(GLA_SV_POST[, "age_band_of_casualty_cas1_under_10"])



# Create new variable for age_band_casualty_11_20
new_var_name <- "age_band_of_casualty_cas1_11_20"

# Create the new variable using ifelse
GLA_SV_POST[new_var_name] <- rowSums(GLA_SV_POST[, c("age_band_of_casualty_cas1_11_15", "age_band_of_casualty_cas1_16_20")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_POST[, "age_band_of_casualty_cas1_11_20"] <- as.numeric(GLA_SV_POST[, "age_band_of_casualty_cas1_11_20"])



# Create new variable for age_band_casualty_over_65
new_var_name <- "age_band_of_casualty_cas1_over_65"

# Create the new variable using ifelse
GLA_SV_POST[new_var_name] <- rowSums(GLA_SV_POST[, c("age_band_of_casualty_cas1_66_75", "age_band_of_casualty_cas1_over_75")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_POST[, "age_band_of_casualty_cas1_over_65"] <- as.numeric(GLA_SV_POST[, "age_band_of_casualty_cas1_over_65"])



# Create new variable for age_band_of_driver_veh1_over_65
new_var_name <- "age_band_of_driver_veh1_over_65"

# Create the new variable using ifelse
GLA_SV_POST[new_var_name] <- rowSums(GLA_SV_POST[, c("age_band_of_driver_veh1_66_75", "age_band_of_driver_veh1_over_75")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_POST[, "age_band_of_driver_veh1_over_65"] <- as.numeric(GLA_SV_POST[, "age_band_of_driver_veh1_over_65"])



# Create new variable for age_band_of_driver_veh1_16_25
new_var_name <- "age_band_of_driver_veh1_16_25"

# Create the new variable using ifelse
GLA_SV_POST[new_var_name] <- rowSums(GLA_SV_POST[, c("age_band_of_driver_veh1_16_20", "age_band_of_driver_veh1_21_25")]) >= 1

# Convert logical values to numeric (1 for TRUE, 0 for FALSE)
GLA_SV_POST[, "age_band_of_driver_veh1_16_25"] <- as.numeric(GLA_SV_POST[, "age_band_of_driver_veh1_16_25"])


```


## 2.3 Correlation matrix for potential IVs in GLA_SV_POST models
```{r, warning=FALSE}

# To control for multicollinearity between potential independent variables, those with pairwise correlation coefficients >0.3 (or <-0.3) were considered too highly correlated to be included in the same model

# Create a correlation matrix for all variables in GLA_SV_POST
cor_matrix <- cor(GLA_SV_POST)

# Write csv to inspect correlation matrix in a more intuitive environment; note, it is often beneficial to apply coloured conditional formatting in Excel to identify variables that exceed 0.3 or -0.3
write.csv(x=cor_matrix, file="Cor_GLA_SV_POST.csv")

```


## 2.4 Recoding, reordering & examining the distribution of the GLA_SV_POST dependent variable (accident severity)
```{r}

# Prior to modelling, the DV ('accident_severity') had to be recoded as a factor; further, the order of 'accident_severity' was inverted so that 1=slight, 2=serious and 3=fatal

# recode DV (using 'car' package) so that 3 is highest outcome, i.e., most severe accidents
GLA_SV_POST$accident_severity <- car::recode(GLA_SV_POST$accident_severity, as.factor=TRUE, "3=1; 2=2; 1=3") 

# Check distribution of DV - 3 should have the fewest outcomes, 1 the most - i.e., heavily zero-inflated distribution
table(GLA_SV_POST$accident_severity) 

# Check % of fatal RTCs (should be 1% or more for ordered modelling, or above 20 observations)
fatal <- (24/1373)*100
fatal

# Note: the % of fatal RTCs is 1.7% (to nearest whole number) and >20; hence, ordered models are appropriate for GLA_SV_POST

```


## 2.5 Stepwise regression for GLA_SV_POST using the 'MASS' package
```{r}

# A stepwise regression (forward selection) algorithm was used to select the starting variables in the fixed parameters ordered probit (FPOP) framework

# Specify model start point with constant (1) only
FitStart = polr(accident_severity ~ 1, data=GLA_SV_POST, na.action = na.omit, method = "probit") 

# Specify potential full model for stepwise regression (checked against correlation matrix; also some variables, including propulsion code, junction control and engine capacity were removed due to sparsity or lack of variation)
FitAll = polr(accident_severity ~ covid + 
                speed_limit_20 + 
                casualty_type_cas1_pedestrian + 
                pedestrian_location_cas1_footway_verge + 
                pedestrian_location_cas1_in_carriageway_not_crossing + 
                pedestrian_location_cas1_not_using_crossing + 
                first_road_class_A + 
                first_road_class_B + 
                road_type_one_way + 
                road_type_dual_c +
                junction_detail_crossroads + 
                pedestrian_crossing_physical_facilities_pelican_puffin +
                pedestrian_crossing_physical_facilities_pedestrian_phase + 
                light_conditions_dark_lit + 
                weather_conditions_rain_no_wind + 
                vehicle_manoeuvre_veh1_reversing + 
                vehicle_manoeuvre_veh1_slowing_stopping + 
                vehicle_manoeuvre_veh1_going_ahead_other + 
                first_point_of_impact_veh1_back + 
                first_point_of_impact_veh1_offside + 
                first_point_of_impact_veh1_no_impact + 
                journey_purpose_of_driver_veh1_commute + 
                sex_of_driver_veh1_male + 
                age_band_of_driver_veh1_16_25 + 
                age_band_of_driver_veh1_over_65 + 
                age_band_of_driver_veh1_over_75 + 
                sex_of_casualty_cas1_male + 
                age_band_of_casualty_cas1_under_10 + 
                age_band_of_casualty_cas1_11_20 + 
                age_band_of_casualty_cas1_36_45 + 
                age_band_of_casualty_cas1_over_65, data=GLA_SV_POST, na.action = na.omit, method = "probit")


# Forward selection algorithm (iterative process based on largest reduction in AIC per potential IV added to model formula)
step(FitStart, direction="forward", scope=formula(FitAll)) 

```


## 2.6 The specification of more complex models (e.g. RPOP or RPOPHM) using the 'Rchoice' package, for GLA_SV_POST data
```{r}

# The 'Rchoice' package allows the specification of more complex models (i.e., RPOP and RPOPHM)

# Re-estimate basic FPOP model (i.e., FPOP_GLA_SV_POST) based on the output of the previous chunk (i.e., the stepwise regression)
# Note only statistically significant IVs are retained: Pr>|z| (i.e., p-value) < 0.05 (95%) or <0.1 (90%)
FPOP_GLA_SV_POST <- Rchoice(accident_severity ~ covid + 
                              age_band_of_casualty_cas1_over_65 + 
                              first_point_of_impact_veh1_back + 
                              first_point_of_impact_veh1_offside + 
                              first_road_class_A + 
                              age_band_of_driver_veh1_over_65 + 
                              sex_of_driver_veh1_male + 
                              casualty_type_cas1_pedestrian + 
                              junction_detail_crossroads + 
                              age_band_of_driver_veh1_16_25 + 
                              sex_of_casualty_cas1_male, data=GLA_SV_POST, family=ordinal("probit"))

# Summarise FPOP model
summary(FPOP_GLA_SV_POST)

```

## 2.6 continued: random parameter trials
```{r}

# Now trial random parameters (using the 'Trial_RP_GLA_SV_POST' formula); note: this is a trial and error process, i.e., try every potential IV (that meets the previously defined inclusion criteria as random parameters)

# Trial RPOP 
Trial_RP_GLA_SV_POST <- Rchoice(accident_severity ~ 
                                  covid + 
                                  age_band_of_casualty_cas1_over_65 + 
                                  first_point_of_impact_veh1_back + 
                                  first_point_of_impact_veh1_offside + 
                                  first_road_class_A + 
                                  age_band_of_driver_veh1_over_65 + 
                                  sex_of_driver_veh1_male + 
                                  casualty_type_cas1_pedestrian + 
                                  junction_detail_crossroads +
                                  age_band_of_driver_veh1_16_25 + 
                                  sex_of_casualty_cas1_male + 
                                  speed_limit_20 + 
                                  light_conditions_dark_lit, data=GLA_SV_POST, family=ordinal("probit"),
                                  ranp=c(age_band_of_casualty_cas1_over_65="n", # specify random parameters (ranp); n = normal distribution
                                  sex_of_driver_veh1_male="n", 
                                  speed_limit_20="n", 
                                  light_conditions_dark_lit="n"), 
                                  R=200) # R = number of Halton draws - 200 is sufficient for ranp trials

# build on FPOP model - trial all available independent variables as random parameters (around 6/7 variables specified as RPs at a time to reduce model estimation time



# Finalised RPOP model
RP_GLA_SV_POST <- Rchoice(accident_severity ~ covid + 
                            age_band_of_casualty_cas1_over_65 + 
                            first_point_of_impact_veh1_back + 
                            first_point_of_impact_veh1_offside + 
                            first_road_class_A + 
                            age_band_of_driver_veh1_over_65 + 
                            sex_of_driver_veh1_male + 
                            casualty_type_cas1_pedestrian + 
                            junction_detail_crossroads + 
                            age_band_of_driver_veh1_16_25 + 
                            sex_of_casualty_cas1_male + 
                            speed_limit_20 + 
                            light_conditions_dark_lit, data=GLA_SV_POST, family=ordinal("probit"), 
                            ranp=c(age_band_of_casualty_cas1_over_65 = "n",
                            sex_of_driver_veh1_male="n", 
                            speed_limit_20="n", 
                            light_conditions_dark_lit="n"), R=500) # R = number of Halton draws - 500 required for final model

# Summarise RPOP model
summary(RP_GLA_SV_POST)


```


## 2.6 continued: re-estimate FPOP with RP variables (for purposes of fair comparison of FPOP and RPOP in model evaluation)
```{r}

# Finalised FPOP model
FP_GLA_SV_POST <- Rchoice(accident_severity ~ covid + 
                            age_band_of_casualty_cas1_over_65 + 
                            first_point_of_impact_veh1_back + 
                            first_point_of_impact_veh1_offside + 
                            first_road_class_A + 
                            age_band_of_driver_veh1_over_65 +
                            sex_of_driver_veh1_male + 
                            casualty_type_cas1_pedestrian + 
                            junction_detail_crossroads + 
                            age_band_of_driver_veh1_16_25 + 
                            sex_of_casualty_cas1_male + 
                            speed_limit_20 + 
                            light_conditions_dark_lit, data=GLA_SV_POST, family=ordinal("probit"))

# Summarise FPOP model 
summary(FP_GLA_SV_POST)

```


## 2.6 continued: test for heterogeneity in the means of random parameter variables (i.e., RPOPHM trials)
```{r}

# Random Parameters Ordered Probit with Heterogeneity in the Means of RPs (RPOPHM) trials can be conducted as shown in 'Edinburgh_models.Rmd'; In the case of GLA_SV_POST, the RPOPHM framework produced statistically significant results (i.e., exogenous variables explained heterogeneity in the means of random parameter variables)

# Finalised RPOPHM model
RPHM_GLA_SV_POST <- Rchoice(accident_severity ~ covid +
                              age_band_of_casualty_cas1_over_65 + 
                              first_point_of_impact_veh1_back + 
                              first_point_of_impact_veh1_offside + 
                              first_road_class_A + 
                              age_band_of_driver_veh1_over_65 + 
                              sex_of_driver_veh1_male + 
                              casualty_type_cas1_pedestrian + 
                              junction_detail_crossroads + 
                              age_band_of_driver_veh1_16_25 + 
                              sex_of_casualty_cas1_male + 
                              speed_limit_20 + 
                              light_conditions_dark_lit | age_band_of_casualty_cas1_over_65 + 
                              sex_of_driver_veh1_male +
                              speed_limit_20 + 
                              light_conditions_dark_lit + 
                              pedestrian_location_cas1_footway_verge, data=GLA_SV_POST, family=ordinal("probit"),
                              ranp=c(age_band_of_casualty_cas1_over_65 = "n", # specify random parameters (ranp); n = normal distribution
                              sex_of_driver_veh1_male="n", 
                              speed_limit_20="n",
                              light_conditions_dark_lit="n"), 
                              R=500, # R = number of Halton draws - 500 required for final model
                              mvar=list(light_conditions_dark_lit=c("pedestrian_location_cas1_footway_verge"))) # Specification of HM in RP variable(s)

# Summarise RPOPHM model
summary(RPHM_GLA_SV_POST)

```


## 2.7: Model evaluation
```{r}

# Likelihood ratio tests are utilised in this analysis (using 'lmtest') to compare the final RPOP model (in this case RP_GLA_SV_POST) to the fixed parameters counterpart (FP_GLA_SV_POST) (LRT I); and also, the final RPOPHM model to the FPOP counterpart (LRT II)

# LRTI: LRT test for RP_GLA_SV_POST versus FP_GLA_SV_POST
LRTI <- lrtest(FP_GLA_SV_POST, RP_GLA_SV_POST)
LRTI

# The LRTI result shows that that RP_GLA_SV_POST is superior to FP_GLA_SV_POST with >99% level of confidence


# LRTII: LRT test for RPHM_GLA_SV_POST versus FP_GLA_SV_POST
LRTII <- lrtest(FP_GLA_SV_POST, RPHM_GLA_SV_POST)
LRTII

# The LRTII result shows that that RPHM_GLA_SV_POST is superior to FP_GLA_SV_POST with >99% level of confidence


```


## -----------------------------------------------------------------------------
## Final model for GLA_SV_POST = RPOPHM (RPHM_GLA_SV_POST)
## -----------------------------------------------------------------------------


